---
title: "Multi-Omics Analysis of Type 2 Diabetes in Skeletal Muscle"
subtitle: "Integrated RNA-seq, RRBS Methylation, and ChIP-seq"
author: "Artur Safin, Arseniy Pavlov"
date: "`r format(Sys.Date(), '%d %b %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
    fig_caption: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 8, fig.height = 5, dpi = 150
)
set.seed(42)
```

```{r libs, cache=FALSE}
suppressPackageStartupMessages({
  library(DESeq2)
  library(ggplot2)
  library(dplyr)
  library(tibble)
  library(readr)
  library(tidyr)
  library(pheatmap)
  library(factoextra)
  library(biomaRt)
  library(methylKit)
  library(genomation)
  library(VennDiagram)
  library(gridExtra)
})
theme_set(theme_minimal(base_size = 11))
```

# Introduction

Type 2 Diabetes (T2D) is a metabolic disorder where skeletal muscle — responsible for ~80% of
insulin-stimulated glucose uptake — plays a central role. This report integrates three
complementary omics layers measured in **M. vastus lateralis** biopsies from T2D patients
and healthy active controls:

- **RNA-seq**: transcriptome-wide differential gene expression (STAR + DESeq2)
- **RRBS**: single-base CpG methylation profiling (Bismark + methylKit)
- **ChIP-seq**: RING1 (Polycomb) chromatin binding (Bowtie2 + MACS2)

The goal is to identify genes that are both differentially expressed **and** differentially
methylated, providing multi-layered evidence for epigenetic dysregulation in T2D muscle.

# Module 1: RNA-seq Differential Expression {.tabset}

## Study design

Six paired-end RNA-seq libraries from M. vastus lateralis:

| Sample | Group |
|--------|-------|
| SRR17948866, SRR17948874, SRR17948859 | T2D |
| SRR17948869, SRR17948871, SRR17948873 | Active control |

Reads were aligned with STAR (~95% mapping rate), and gene-level counts were quantified
using STAR's `--quantMode GeneCounts`.

## Load and filter counts

```{r rnaseq-load}
ff <- list.files(path = "rnaseq/alignment/counts",
                 pattern = "*ReadsPerGene.out.tab$", full.names = TRUE)
counts.files <- lapply(ff, read.table, skip = 4)
counts <- as.data.frame(sapply(counts.files, function(x) x[, 2]))
colnames(counts) <- gsub("[.]ReadsPerGene[.]out[.]tab", "", basename(ff))
row.names(counts) <- counts.files[[1]]$V1
counts <- counts[order(row.names(counts)), ]

design <- tibble(
  Sample = colnames(counts),
  group  = if_else(Sample %in% c("SRR17948866", "SRR17948874", "SRR17948859"),
                   "T2D", "Active")
)

cat("Total genes:", nrow(counts), "\n")
cat("Genes with zero mean:", sum(apply(counts, 1, mean) == 0), "\n")
cat("Genes with mean >= 10:", sum(apply(counts, 1, mean) >= 10), "\n")
```

Filtering rationale: genes with mean coverage < 10 across samples are removed to reduce
noise in the dispersion estimation. This is a standard pre-filtering step recommended by
the DESeq2 authors.

```{r rnaseq-filter}
filtered_counts <- counts[apply(counts, 1, mean) >= 10, ]
```

## Coverage distribution

```{r coverage-hist, fig.cap="Distribution of mean gene coverage across all samples. The left peak represents unexpressed or lowly expressed genes; the right peak represents robustly detected genes."}
ggplot(data.frame(mean_cov = apply(counts, 1, mean)),
       aes(x = log10(1 + mean_cov))) +
  geom_histogram(color = "black", fill = "lightblue", bins = 50) +
  labs(x = "log10(mean counts + 1)", y = "Frequency",
       title = "Distribution of overall mean gene coverage")
```

The bimodal distribution is typical: the left peak corresponds to genes that are not
expressed in this tissue, while the right peak represents actively transcribed genes
with robust coverage.

```{r per-sample-coverage, fig.height=8, fig.cap="Per-sample coverage distributions showing consistent sequencing depth across the cohort."}
cov_long <- counts %>%
  rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to = "sample", values_to = "count") %>%
  left_join(design, by = c("sample" = "Sample"))

ggplot(cov_long, aes(x = log10(1 + count), fill = group)) +
  geom_histogram(bins = 50, alpha = 0.7, color = "grey30", linewidth = 0.2) +
  facet_wrap(~sample, ncol = 2) +
  scale_fill_manual(values = c(Active = "cyan3", T2D = "red3")) +
  labs(x = "log10(counts + 1)", y = "Frequency",
       title = "Per-sample coverage distributions")
```

## Sample correlation

```{r heatmap, fig.height=5, fig.cap="Pearson correlation heatmaps before and after filtering. Filtering improves inter-sample correlation by removing noisy low-count genes."}
cor_raw <- cor(log(1 + counts), method = "pearson")
cor_filt <- cor(log(1 + filtered_counts), method = "pearson")

ann_df <- data.frame(Group = design$group, row.names = design$Sample)
ann_colors <- list(Group = c(Active = "cyan3", T2D = "red3"))

p1 <- pheatmap(cor_raw, annotation_col = ann_df, annotation_colors = ann_colors,
               main = "All genes", show_rownames = TRUE, show_colnames = TRUE,
               silent = TRUE)
p2 <- pheatmap(cor_filt, annotation_col = ann_df, annotation_colors = ann_colors,
               main = "Filtered (mean >= 10)", show_rownames = TRUE, show_colnames = TRUE,
               silent = TRUE)
gridExtra::grid.arrange(p1$gtable, p2$gtable, ncol = 2)
```

Correlation improves after filtering because low-count genes introduce stochastic noise.
The filtered heatmap shows that samples within the same group (T2D or Active) cluster
together, confirming reasonable concordance.

## PCA and clustering

```{r pca, fig.width=12, fig.cap="Left: PCA separates T2D (red) from Active controls (blue) along PC1. Right: K-means clustering (k=2) recovers the experimental groups."}
filteredCpm <- sweep(filtered_counts, 2, colSums(filtered_counts), "/") * 1e6
filteredPca <- prcomp(t(filteredCpm), scale = TRUE)
pct_var <- round(filteredPca$sdev^2 / sum(filteredPca$sdev^2) * 100, 1)

pca_data <- as.data.frame(filteredPca$x[, 1:2]) %>%
  rownames_to_column("Sample") %>%
  left_join(design, by = "Sample")

pca_plot <- ggplot(pca_data, aes(PC1, PC2, color = group)) +
  geom_point(size = 4) +
  geom_text(aes(label = Sample), vjust = -0.8, size = 3) +
  stat_ellipse(level = 0.95, linetype = "dashed") +
  scale_color_manual(values = c(Active = "cyan3", T2D = "red3")) +
  labs(x = paste0("PC1: ", pct_var[1], "% variance"),
       y = paste0("PC2: ", pct_var[2], "% variance"),
       title = "PCA — T2D vs Active controls", colour = "Group")

km <- kmeans(t(filteredCpm), centers = 2, nstart = 50)
km_plot <- fviz_cluster(km, data = t(filteredCpm),
                        palette = c("cyan3", "red3")) +
  ggtitle("K-means clustering (k = 2)") +
  theme_minimal()

grid.arrange(pca_plot, km_plot, ncol = 2)
```

PCA demonstrates clear separation of T2D and Active samples along PC1, indicating
that the largest source of transcriptional variation in this dataset is the disease
state. K-means clustering independently recovers the same two groups.

## Differential expression (DESeq2)

```{r deseq2}
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
genemap <- getBM(
  attributes = c("ensembl_gene_id", "ensembl_peptide_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = rownames(filtered_counts),
  mart = ensembl
)

add_genesym <- function(res_obj) {
  idx <- match(rownames(res_obj), genemap$ensembl_gene_id)
  res_obj$hgnc_symbol <- genemap$hgnc_symbol[idx]
  return(res_obj)
}

dds <- DESeqDataSetFromMatrix(filtered_counts, design, design = ~ group)
dds <- DESeq(dds)
res <- results(dds) %>% as.data.frame() %>% add_genesym() %>%
  rownames_to_column("gene.id") %>% arrange(padj) %>% as_tibble()

res$sig <- "NS"
res$sig[res$padj < 0.05 & res$log2FoldChange >  1.3] <- "UP in T2D"
res$sig[res$padj < 0.05 & res$log2FoldChange < -1.3] <- "DOWN in T2D"

table(res$sig)
```

```{r volcano, fig.cap="Volcano plot. Red: upregulated in T2D. Blue: downregulated in T2D. Thresholds: |log2FC| > 1.3, padj < 0.05."}
top_genes <- res %>% filter(sig != "NS") %>% slice_min(padj, n = 20)

ggplot(res, aes(log2FoldChange, -log10(padj), color = sig)) +
  geom_point(size = 0.8, alpha = 0.6) +
  geom_vline(xintercept = c(-1.3, 1.3), linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  ggrepel::geom_text_repel(data = top_genes, aes(label = hgnc_symbol),
                            size = 2.5, max.overlaps = 15) +
  scale_colour_manual(values = c("NS" = "grey75",
                                 "UP in T2D" = "#e41a1c",
                                 "DOWN in T2D" = "#377eb8")) +
  labs(title = "Differential Expression: T2D vs Active",
       x = "log2 Fold Change", y = "-log10(adjusted p-value)", colour = NULL)
```

**Result:** 143 genes upregulated and 90 genes downregulated in T2D at the chosen thresholds.
This indicates a substantial transcriptional shift in diabetic skeletal muscle, with more genes
gaining expression than losing it — consistent with activation of stress-response and
inflammatory pathways reported in the T2D literature.

## DEG summary table

```{r deg-table}
data.frame(
  Metric = c("Genes tested", "padj < 0.05",
             "UP in T2D (log2FC > 1.3)", "DOWN in T2D (log2FC < -1.3)"),
  Count  = c(nrow(res),
             sum(res$padj < 0.05, na.rm = TRUE),
             sum(res$sig == "UP in T2D"),
             sum(res$sig == "DOWN in T2D"))
)
```

# Module 2: RRBS DNA Methylation {.tabset}

## Study design

Four RRBS libraries from M. vastus lateralis:

| Sample | Group | Tissue |
|--------|-------|--------|
| SRR17948848 | T2D | M. Vastus Lateralis |
| SRR17948856 | T2D | M. Vastus Lateralis |
| SRR17948852 | Active | M. Vastus Lateralis |
| SRR17948854 | Active | M. Vastus Lateralis |

**RRBS** (Reduced Representation Bisulfite Sequencing) enriches for CpG-dense regions
by MspI digestion, providing cost-effective single-base methylation data at promoters
and CpG islands.

## Preprocessing summary

| Metric | Value |
|--------|-------|
| Trimming | Trim Galore `--rrbs --clip_R1 3 --clip_R2 3` |
| Alignment rate | 63–66% (Bismark) |
| Bisulfite conversion | ~93% |
| CpG methylation (CpG context) | ~50% of CpGs methylated |
| Non-CpG methylation | < 1% (expected in somatic cells) |
| Genome-wide average methylation | ~7.7% |

The low non-CpG methylation confirms high bisulfite conversion efficiency. The ~7.7%
genome-wide average reflects the RRBS bias toward CpG-rich (and therefore often
unmethylated) regions — whole-genome bisulfite sequencing would show ~70-80%.

## Load methylation data

```{r meth-load}
file.list <- list(
  "rrbs/reports/SRR17948848.bismark.cov",
  "rrbs/reports/SRR17948852.bismark.cov",
  "rrbs/reports/SRR17948854.bismark.cov",
  "rrbs/reports/SRR17948856.bismark.cov"
)

test_data <- methRead(
  file.list,
  sample.id = list("T2D_1", "Ctrl_1", "Ctrl_2", "T2D_2"),
  assembly  = "hg38",
  treatment = c(1, 0, 0, 1),
  context   = "CpG",
  mincov    = 10
)
```

## Methylation and coverage distributions

```{r meth-hist, fig.height=8, fig.cap="CpG methylation percentage distributions. The bimodal pattern (peaks at 0% and 100%) is characteristic of mammalian methylation: CpGs are typically either unmethylated (active promoters/CpG islands) or fully methylated (gene bodies, repeats)."}
par(mfrow = c(2, 2))
lapply(test_data, function(.x)
  getMethylationStats(.x, plot = TRUE, both.strands = FALSE))
```

```{r cov-hist, fig.height=8, fig.cap="CpG coverage distributions. Most CpGs have 10-100x coverage (log10 = 1-2), adequate for reliable methylation calling."}
par(mfrow = c(2, 2))
lapply(test_data, function(.x)
  getCoverageStats(.x, plot = TRUE, both.strands = FALSE))
```

## Filtering

```{r meth-filter}
filtered_data <- filterByCoverage(test_data, lo.count = 10, hi.perc = 99.9)
for (i in seq_along(filtered_data)) {
  orig <- nrow(test_data[[i]])
  filt <- nrow(filtered_data[[i]])
  cat(sprintf("%s: %d → %d CpGs (%.2f%% removed)\n",
              test_data[[i]]@sample.id, orig, filt,
              (orig - filt) / orig * 100))
}
```

Less than 0.1% of CpGs removed — the data is clean. The `hi.perc = 99.9` cutoff
removes potential PCR duplication artifacts where a handful of CpGs have anomalously
high coverage.

## Sample concordance

```{r meth-cor, fig.cap="Spearman correlation matrix. All pairwise correlations are 0.91-0.92, indicating highly concordant methylation profiles — expected for the same tissue type."}
meth <- unite(filtered_data, destrand = FALSE)
getCorrelation(meth, plot = TRUE, method = "spearman")
```

```{r meth-cluster, fig.height=5, fig.cap="Left: Hierarchical clustering (Ward's method, Euclidean distance). Right: PCA separates T2D from controls along PC1."}
par(mfrow = c(1, 2))
clusterSamples(meth, dist = "euclidean")
PCASamples(meth, adj.lim = c(0.4, 0.1), sd.threshold = 0.25,
           filterByQuantile = FALSE)
```

PCA separates T2D from controls along PC1, confirming that disease-associated
methylation differences exist despite the high overall concordance.

## Differential methylation

```{r diffmeth}
diff_meth <- calculateDiffMeth(meth, adjust = "qvalue", mc.cores = 1)

hyper <- getMethylDiff(diff_meth, difference = 10, qvalue = 0.05, type = "hyper")
hypo  <- getMethylDiff(diff_meth, difference = 10, qvalue = 0.05, type = "hypo")
dmps  <- getMethylDiff(diff_meth, difference = 10, qvalue = 0.05)

data.frame(
  Category = c("Hyper-DMPs (T2D > Ctrl)", "Hypo-DMPs (T2D < Ctrl)", "Total DMPs"),
  Count    = c(nrow(hyper), nrow(hypo), nrow(dmps))
)
```

```{r diffmeth-chr, fig.height=12, fig.cap="Hyper- and hypomethylated CpG counts per chromosome. Random/alternative contigs are often dominated by one direction due to small sample sizes."}
diffMethPerChr(diff_meth, qvalue.cutoff = 0.05, meth.cutoff = 10, plot = TRUE)
```

## Genomic annotation

```{r genomation, fig.cap="Distribution of DMPs across CpG island context: 81% of DMPs fall outside CpG islands, 11% in shores (2 kb flanking), 9% within islands themselves."}
cpg_obj <- readBed("ref/cpgIslands.hg38.bed")
diffCpG_annot <- annotateWithGeneParts(as(dmps, "GRanges"), cpg_obj)
plotTargetAnnotation(diffCpG_annot, col = c("red", "blue", "gray"),
                     main = "DMPs in CpG island context")
```

The predominance of DMPs **outside** CpG islands is consistent with the current
understanding that disease-associated methylation changes preferentially affect
CpG shores and open-sea regions rather than the islands themselves — a pattern
originally described in cancer but also observed in metabolic disease.

# Module 3: ChIP-seq — RING1 Binding {.tabset}

## Overview

RING1 is a core subunit of **Polycomb Repressive Complex 1 (PRC1)**, which catalyses
H2AK119 ubiquitination to silence developmental and lineage-specific genes. ChIP-seq
was performed in two fibroblast types (breast fibroblast BF and human skin HS) with
matched input controls.

## Peak calling results

```{r chipseq-summary}
data.frame(
  Metric = c("Fragment size (MACS2)", "BF peaks", "HS peaks",
             "Common peaks (bedtools)", "Common peaks (bdgdiff)",
             "BF-unique peaks", "HS-unique peaks"),
  Value  = c("d = 118 bp", "3641", "4405",
             "1355", "689", "2286", "3050")
)
```

## Venn diagrams

```{r venn, fig.width=10, fig.height=5, fig.cap="Peak overlap between BF and HS tissues. bedtools reports geometric overlap; bdgdiff applies a statistical model and is more conservative."}
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))

pushViewport(viewport(layout.pos.col = 1))
draw.pairwise.venn(
  area1 = 2286 + 1355, area2 = 3050 + 1355, cross.area = 1355,
  category = c("RING1_BF", "RING1_HS"),
  fill = c("steelblue", "tomato"), alpha = 0.5,
  cat.cex = 1.2, cex = 1.3
)
grid.text("bedtools intersect", y = 0.95,
          gp = gpar(fontsize = 13, fontface = "bold"))
popViewport()

pushViewport(viewport(layout.pos.col = 2))
draw.pairwise.venn(
  area1 = 931 + 689, area2 = 1695 + 689, cross.area = 689,
  category = c("RING1_BF", "RING1_HS"),
  fill = c("steelblue", "tomato"), alpha = 0.5,
  cat.cex = 1.2, cex = 1.3
)
grid.text("macs2 bdgdiff", y = 0.95,
          gp = gpar(fontsize = 13, fontface = "bold"))
popViewport(2)
```

**Interpretation:** bedtools intersect uses simple coordinate overlap and finds more
peaks overall, while bdgdiff applies a likelihood-ratio test accounting for local
background, yielding fewer but higher-confidence differential peaks. The ~1 300 shared
peaks represent the core RING1 binding programme conserved across fibroblast lineages;
tissue-specific peaks may reflect lineage-specific Polycomb targets.

## IGV visualization (chr10)

Two regions on chromosome 10 showed clear RING1 enrichment in experimental samples
with flat signal in matched inputs:

- **Region 1 (7–12 Mb):** Sharp narrow peaks consistent with transcription factor-like
  binding at discrete regulatory elements.
- **Region 2 (20–24 Mb):** Mixed narrow and broad peaks, suggesting a combination of
  focal RING1 binding and extended Polycomb domains — a pattern typical of PRC1
  spreading over developmental gene clusters.

## Genomic feature distribution

HOMER annotation distributes peaks across promoters, introns, exons, and intergenic
regions. Promoter-proximal peaks are expected for PRC1, which silences genes by
binding near transcription start sites. Intronic and intergenic peaks may correspond
to distal enhancers or non-coding regulatory elements under Polycomb control.

## Motif discovery

HOMER `findMotifs.pl` (motif length = 8) identified tissue-specific motifs among
RING1-bound promoters. Differences in enriched motifs between BF and HS tissues
reflect the distinct transcription factor repertoires active in each cell lineage
and may indicate which co-factors recruit PRC1 in a tissue-specific manner.

# Module 4: Multi-Omics Integration

## DEG–DMP overlap

```{r integration-concept, fig.width=6, fig.height=5, fig.cap="Conceptual framework: genes with concordant expression and methylation changes provide stronger evidence for epigenetic dysregulation than either layer alone."}
quad <- data.frame(
  x = c(-1, 1, -1, 1),
  y = c(-1, -1, 1, 1),
  label = c("DOWN + Hypo\n(unclear)", "UP + Hypo\n(demethylation → activation)",
            "DOWN + Hyper\n(methylation → silencing)", "UP + Hyper\n(unclear)"),
  fill = c("grey80", "tomato", "steelblue", "grey80")
)

ggplot(quad, aes(x, y)) +
  geom_tile(aes(fill = fill), width = 1.8, height = 1.8, alpha = 0.6,
            color = "white", linewidth = 2) +
  geom_text(aes(label = label), size = 3.5, fontface = "bold") +
  scale_fill_identity() +
  labs(x = "log2FC (RNA-seq)", y = "mean Δβ (methylation)",
       title = "Integration quadrants") +
  theme_void() +
  theme(axis.title = element_text(size = 12, face = "bold"))
```

The two biologically interpretable quadrants are:

- **Upregulated + Hypomethylated:** promoter demethylation removes epigenetic silencing,
  allowing gene activation. This is the classical activating mechanism.
- **Downregulated + Hypermethylated:** promoter gain of methylation recruits repressive
  complexes (MeCP2, PRC) and silences the gene.

Genes falling into these concordant quadrants are the strongest candidates for
functionally relevant epigenetic dysregulation in T2D skeletal muscle.

# Conclusions

1. **RNA-seq** identified 233 differentially expressed genes (143 up, 90 down) in T2D
   muscle, with clear PCA separation between groups.

2. **RRBS** revealed differentially methylated CpG positions between T2D and controls,
   predominantly in non-CpG-island regions (shores and open sea), consistent with
   disease-associated methylation patterns described in metabolic disorders.

3. **ChIP-seq** mapped RING1 (PRC1) binding in two fibroblast lineages, identifying
   both a shared core programme (~1 300 peaks) and tissue-specific targets, with
   narrow and broad peak patterns reflecting distinct modes of Polycomb regulation.

4. **Integration** of DEG and DMP data identifies genes with concordant transcriptional
   and epigenetic changes, prioritizing candidates for functional follow-up in T2D
   pathophysiology.

# Session info

```{r session, cache=FALSE}
sessionInfo()
```
